### 2025-09-05 YEAR2/AER210_LectureNotes/main.tex
sept 5

>>>>

% set section counter from 15 
\setcounter{section}{14}
\paragraph{Note} the section numbering is based on Stewart's book.
\section{Double and Triple Integrals}
\begin{definition}[Double Integral]
    Let $f(x,y)$ be a function defined on a closed and bounded region $R$ in the $xy$-plane. The double integral of $f$ over $R$ is denoted by
    \begin{equation}
        \iint_R f(x,y) \, dA = \iint_R f(x,y) \, dA
    \end{equation}
    where $dA$ represents an infinitesimal area element in the region $R$. The double integral can be interpreted as the volume under the surface defined by $z = f(x,y)$ over the region $R$.
\end{definition}

\subsection{Double Integrals in a Rectangular Region}

 By the point of seeing this note, you should be familiar with the simple case of rectangular, simple cases are provided as examples:

\begin{example}
    Find the area under the quadric surface $z = 16 - x^2 - y^2$ over the square region $R = \{ (x,y) \mid 0 \le x \le 2, 0 \le y \le 2\}$.

    \textbf{Note} We would have to ensure that the surface is above the $xy$-plane in the region of interest, which is true in this case.

    We can set up the double integral as follows:
    $$
        \iint_R (16 - x^2 - y^2) \, dA = \int_0^2 \int_0^2 (16 - x^2
        - y^2) \, dy \, dx
    $$
    First, we integrate with respect to $y$:
    $$
        \int_0^2 (16 - x^2 - y^2) \, dy = \left[ 16y - x^2y - \frac{y^3}{3} \right]_0^2 = 32 - 2x^2 - \frac{8}{3} = \frac{88}{3} - 2x^2
    $$
    Next, we integrate with respect to $x$:
    $$
        \int_0^2 \left( \frac{88}{3} - 2x^2 \right) \, dx = \left[ \frac{88}{3}x - \frac{2x^3}{3} \right]_0^2 = \frac{176}{3} - \frac{16}{3} = \frac{160}{3}
    $$
    Therefore, the area under the surface is $\frac{160}{3}$.

\end{example}

\begin{example}
    Evaluate the double integral of $f(x,y) = x - 3y^2$ over the rectangular region $R = \{ (x,y) \mid 0 \le x \le 2, 1 \le y \le 2\}$.

    We can set up the double integral as follows:
    $$
        \iint_R (x - 3y^2) \, dA = \int_0^2 \int_1^2 (x - 3y^2) \, dy \, dx
    $$
    First, we integrate with respect to $y$:
    $$
        \int_1^2 (x - 3y^2) \, dy = \left[ xy - y^3 \right]_1^2 = 2x - 8 - (x - 1) = x - 7
    $$
    Next, we integrate with respect to $x$:
    $$
        \int_0^2 (x - 7) \, dx = \left[ \frac{x^2}{2} - 7x \right]_0^2 = \left( 2 - 14 \right) - 0 = -12
    $$
    Therefore, the value of the double integral is $-12$.
\end{example}

\begin{theorem}
    If the integrand function $f(x,y)$ is seperable, i.e., $f(x,y) = g(x)h(y)$, then the double integral can be computed as follows:
    \begin{equation}
        \iint_R f(x,y) \, dA = \left( \int_a^b g(x) \, dx \right) \left( \int_c^d h(y) \, dy \right)
    \end{equation}
    where $R = \{ (x,y) \mid a \le x \le b, c \le y \le d \}$.
\end{theorem}
\begin{proof}
    \textbf{Sketch:} $h(y)$ is a constant when integrating with respect to $x$, and vice versa.
\end{proof}

\begin{example}
    Let $f(x,y) = \sin{x} \cos{y}$ and $R = \{ (x,y) \mid 0 \le x \le \frac{\pi}{2}, 0 \le y \le \frac{\pi}{2} \}$. Evaluate the double integral $\iint_R f(x,y) \, dA$.

    Since $f(x,y)$ is seperable, we can write:
    $$
        \iint_R f(x,y) \, dA = \left( \int_0^{\frac{\pi}{2}} \sin{x} \, dx \right) \left( \int_0^{\frac{\pi}{2}} \cos{y} \, dy \right)
    $$
    Evaluating each integral separately gives 1 for both, so the final result is: $1 \times 1 = 1$.
\end{example}
\subsection{Double Integrals in General Regions}
\paragraph{Types of Regions} When the region $R$ is not rectangular, we can still compute the double integral by expressing the region in terms of inequalities. There are three common types of regions:
\begin{definition}[Type I Region]
    A region $R$ is called a Type I region if it can be described by the inequalities:
    $$
        a \le x \le b, \quad g_1(x) \le y \le g_2(x)
    $$
    where $g_1(x)$ and $g_2(x)$ are continuous functions on the interval $[a,b]$. Then, to evaluate the double integral over a Type I region for a continuous function $f(x,y)$, we set up the integral as follows:

    \begin{equation}
        \iint_R f(x,y) \, dA = \int_a^b \int_{g_1(x)}^{g_2(x)} f(x,y) \, dy \, dx
    \end{equation}

    \textbf{Integral Order:} Integrate with respect to $y$ first, then $x$.

    \textbf{Intuition:} As we traverse the outer part ($x$), we are summing up vertical slices (in $y$), and the bounds of those slices depend on $x$ and changes.
\end{definition}

\begin{definition}[Type II Region]
    Type II reigion is similar to Type I, but the roles of $x$ and $y$ are swapped. A region $R$ is called a Type II region if it can be described by the inequalities:
    $$
        c \le y \le d, \quad h_1(y) \le x \le h_2(y)
    $$
    where $h_1(y)$ and $h_2(y)$ are continuous functions on the interval $[c,d]$. Then, to evaluate the double integral over a Type II region for a continuous function $f(x,y)$, we set up the integral as follows: 
    \begin{equation}
        \iint_R f(x,y) \, dA = \int_c^d \int_{h_1(y)}^{h_2(y)} f(x,y) \, dx \, dy
    \end{equation}

    The integral order and intuition is mirrored from Type I, but we are summing up horizontal slices (in $x$), and the bounds of those slices depend on $y$ and changes.
\end{definition}

\begin{definition}[Type III Region]
    A region $R$ is called a Type III region if it can be described as the union of a finite number of Type I and Type II regions. To evaluate the double integral over a Type III region for a continuous function $f(x,y)$, we can break down the integral into separate integrals over each Type I or Type II subregion and sum them up:
    \begin{equation}
        \iint_R f(x,y) \, dA = \sum_{i=1}^n \iint_{R_i} f(x,y) \, dA
    \end{equation}
    where each $R_i$ is either a Type I or Type II region. And that:
    $$
        \bigcup_{i=1}^n R_i = R \quad \text{and} \quad R_i \cap R_j = \emptyset \text{ for } i \neq j
    $$
    This approach allows us to handle more complex regions by breaking them down into simpler parts.
\end{definition}

\begin{example}
    Find the volume of the solid that lies under the paraboloid $z = f(x,y) = x^2 + y^2$ and above the region $R$ bounded
     by $y = 2x$ and $y = x^2$.

     First, you would sketch the region to understand its shape and boundaries at Figure \ref{fig:region}.

    \begin{figure}[h]
        \centering
        \begin{tikzpicture}
            \begin{axis}[
                axis lines = middle,
                xlabel = $x$,
                ylabel = $y$,
                xmin = -1, xmax = 3,
                ymin = -1, ymax = 5,
                xtick = {0,1,2,3},
                ytick = {0,1,2,3,4},
                grid = both,
                minor tick num = 1,
                domain = -1:3,
            ]
            \addplot[blue, thick, name path=twox, name path global=twox] {2*x};
            \addplot[red, thick, name path=xsq, name path global=xsq] {x^2};
            \addplot[fill=gray, opacity=0.5] fill between[of=twox and xsq, soft clip={domain=0:2}];          \end{axis}
        \end{tikzpicture}
        \caption{Region bounded by $y = 2x$ and $y = x^2$} \label{fig:region}
    \end{figure}

    We can tell that this is a Type I regionwhere $0 \le x \le 2$, and $x^2 \le y \le 2x$. Thus, we can set up the double integral as follows:
    $$
        \iint_R (x^2 + y^2) \, dA = \int_0^2 \int_{x^2}^{2x} (x^2 + y^2) \, dy \, dx
    $$
    First, we integrate with respect to $y$:
    $$
        \int_{x^2}^{2x} (x^2 + y^2) \, dy = \left[ x^2y + \frac{y^3}{3} \right]_{y=x^2}^{y=2x} = 2x^3 + \frac{8x^3}{3} - x^4 - \frac{x^6}{3} = \frac{14x^3}{3} - x^4 - \frac{x^6}{3}
    $$
    Next, we integrate with respect to $x$:
    $$
        \int_0^2 \left( \frac{14x^3}{3} - x^4 - \frac{x^6}{3} \right) \, dx = \left[ \frac{14x^4}{12} - \frac{x^5}{5} - \frac{x^7}{21} \right]_0^2 = \frac{216}{35}
    $$
    Therefore, the volume of the solid is $\frac{216}{35}$.

\end{example}

\begin{example}
    Consider the above example, but we want to set it up as a Type II region. The region $R$ can be described by $0 \le y \le 4$, and $\frac{y}{2} \le x \le \sqrt{y}$. Thus, we can set up the double integral as follows:
    $$
        \iint_R (x^2 + y^2) \, dA = \int_0^4 \int_{\frac{y}{2}}^{\sqrt{y}} (x^2 + y^2) \, dx \, dy
    $$
    First, we integrate with respect to $x$:
    $$
        \int_{\frac{y}{2}}^{\sqrt{y}} (x^2 + y^2) \, dx = \left[ \frac{x^3}{3} + y^2x \right]_{x=\frac{y}{2}}^{x=\sqrt{y}} = \frac{y^{\frac{3}{2}}}{3
        } + y^{\frac{5}{2}} - \frac{y^3}{24} - \frac{y^3}{2} = \frac{y^{\frac{3}{2}}}{3} + y^{\frac{5}{2}} - \frac{13y^3}{24}
    $$
    Next, we integrate with respect to $y$:
    $$
        \int_0^4 \left( \frac{y^{\frac{3}{2}}}{3} + y^{\frac{5}{2}} - \frac{13y^3}{24} \right) \, dy = \left[ \frac{2y^{\frac{5}{2}}}{15} + \frac{2y^{\frac{7}{2}}}{7} - \frac{13y^4}{96} \right]_0^4 = \frac{216}{35}
    $$
    Therefore, the volume of the solid is $\frac{216}{35}$, which is consistent with the previous result. This is also consistent with Fubini's Theorem.
\end{example}

\begin{example}
    Integrate the surface given by $z = e^{x^2}$ over the triangular region with vertices at $(0,0)$, $(1,0)$, and $(1,1)$. We can describe the region as either a Type I or Type II region:

    (Incorrect) Here, we will describe it as a Type II regionwhere $0 \le y \le 1$, and $y \le x \le 1$. Thus, we can set up the double integral as follows:
    $$
        \iint_R e^{x^2} \, dA = \int_0^1 \int_y^1 e^{x^2} \, dx \, dy
    $$
    We can tell that $e^{x^2}$ does not have an elementary antiderivative, so we cannot integrate with respect to $x$ directly.

    (Correct) However, we can change the order of integration to make it a Type I regionwhere $0 \le x \le 1$, and $0 \le y \le x$. Thus, we can set up the double integral as follows:
    $$
        \iint_R e^{x^2} \, dA = \int_0^1 \int_0^x e^{x^2} \, dy \, dx
    $$
    First, we integrate with respect to $y$:
    $$
        \int_0^x e^{x^2} \, dy = \left[ y e^{x^2} \right]_0^x = xe^{x^2}
    $$
    Next, we integrate with respect to $x$:
    $$
        \int_0^1 xe^{x^2} \, dx
    $$
    This is now obvious, a simple $u$-substitution with $u = x^2$, $du = 2x \, dx$:
    $$
        \int_0^1 xe^{x^2} \, dx = \frac{1}{2} \int_0^1 e^u \, du = \frac{1}{2} \left[ e^u \right]_0^1 = \frac{e - 1}{2}
    $$
    Therefore, the value of the double integral is $\frac{e - 1}{2}$.

    \textbf{Intuition} When the integrand is difficult to integrate with respect to one variable, consider changing the order of integration. You should be able to tell that $e^{x^2}$ has no elementary antiderivative, so you would have ruled out integrating with respect to $x$ first.
\end{example}

\subsection{Formal Definition of Double Integrals}
There is two definitions of double integrals in this course, due to the discrepancy between Stewart's book and the lectures.

\begin{shaded}
    \subsubsection*{\textbf{Review.  } Formal Definition of Definite Integral (Single Variable)}

    Consider $y = f(x) \ge 0$ on the interval $x \in [a,b]$. We divide the interval into $n$ subintervals of equal width $\Delta x = \frac{b-a}{n}$, and let $x_i^*$ be a sample point in the $i$-th subinterval. The Riemann sum is given by:
    $$
        S_n = \sum_{i=1}^n f(x_i^*) \Delta x
    $$
    Now, for any $x_i^*$, we consider the minimum and maximum values of $f(x_i^*)$ in the $i$-th subinterval, denoted as $m_i$ and $M_i$ respectively. We can then define the lower sum $L_n$ and upper sum $U_n$ as follows:
    $$
        L_n = \sum_{i=1}^n m_i \Delta x \quad \text{and} \quad U_n = \sum_{i=1}^n M_i \Delta x
    $$
    To satisfy the squeeze theorem, for all $i$, we would need:
    $$
        \lim_{n \to \infty} M_i - m_i = \lim_{\delta x \to 0} M_i - m_i = 0
    $$
    If $f(x)$ is continuous on $[a,b]$. Then, we have:
    $$
        \lim_{n \to \infty} L_n = \lim_{n \to \infty} U_n = \int_a^b f(x) \, dx
    $$

    For the case of discontinuous functions, if the set of discontinuities has measure zero, then the function is still integrable.
\end{shaded}

\begin{definition}[Definition of Double Integral]
    Let $R$ be a rectangular region in the $xy$-plane given by $R = [a,b] \times [c,d]$. The double integral of a function $f(x,y)$ over the region $R$ is defined as:
    \begin{subequations}
    \begin{equation}
        \iint_R f(x,y) \, dA = \lim_{n \to \infty} \sum_{i=1}^n f(x_i^*,y_i^*) \Delta A_i \quad (\text{Riemann Definition})
    \end{equation}
    where $\Delta A_i$ is the area of the $i$-th subrectangle, and $(x_i^*,y_i^*)$ is a sample point in it. The limit is taken as the maximum diameter of the subrectangles approaches zero.
    \begin{equation}
        \iint_R f(x,y) \, dA = \lim_{n,m \to \infty} \sum_{i=1}^n \sum_{j=1}^m f(x_i^*,y_j^*) \Delta A_{ij} \quad (\text{Grid Formulation})
    \end{equation}
    where $\Delta A_{ij}$ is the area of the $ij$-th subrectangle, and $(x_i^*,y_j^*)$ is a sample point in it. Note that the $\Delta A_{ij}$ may be non-uniform. The limit is taken as the maximum diameter of the subrectangles approaches zero.
    \end{subequations}

    Similarly, the lower and upper sums for double integrals are:
    \begin{subequations}
    \begin{equation}
        L_n = \sum_{i=1}^n m_i \Delta A_i \quad \text{and} \quad U_n = \sum_{i=1}^n M_i \Delta A_i \quad (\text{Riemann Definition})
    \end{equation}
    \begin{equation}
        L_{n,m} = \sum_{i=1}^n \sum_{j=1}^m m_{ij} \Delta A_{ij} \quad \text{and} \quad U_{n,m} = \sum_{i=1}^n \sum_{j=1}^m M_{ij} \Delta A_{ij} \quad (\text{Grid Formulation})
    \end{equation}
    \end{subequations}
    Here, $m_{ij}$ and $M_{ij}$ are the minimum and maximum values of $f(x,y)$ in the $ij$-th subrectangle. Define $||P|| = \max ||(\Delta x_i, \Delta y_j)||$ as the maximum diameter of the subrectangles. For the squeeze theorem, we require:
    $$
        \lim_{n,m \to \infty} (M_{ij} - m_{ij}) = \lim_{||P|| \to 0} (M_{ij} - m_{ij}) = 0
    $$
    If $f(x,y)$ is continuous on $R$, then:
    $$
        \lim_{n,m \to \infty} L_{n,m} = \lim_{n,m \to \infty} U_{n,m} = \iint_R f(x,y) \, dA
    $$

    The Riemann definition and grid formulation are similar.
\end{definition}

### 2025-09-05 YEAR2/ECE253_LectureNotes/main.tex
4 sept
sept 5

\documentclass[11pt]{article}
\input{preamble}
\begin{document}
\thispagestyle{empty}
{\LARGE \bf ECE 253 Lecture Notes}\\
{\large Hei Shing Cheung}\\
Digital and Computer Systems, Fall 2025 \hfill ECE253\\
\\
The up-to-date version of this document can be found at \url{https://github.com/HaysonC/skulenotes}\\

\begin{shaded}
\section*{Introduction}
\paragraph{Layers of Computation} In hardware, we have the following layers of abstraction:
\begin{itemize}
    \item Computation
    \item Adders 
    \item Logic Gates
    \item Transistors
    \item Silicon
\end{itemize}
In this course, we will focus on the first three layers, on top of the logic gate level.

\paragraph{Layer of abstraction} At this course, for the digital systems part, we would start from understanding logic gates, all the way to understanding computer architecture, with each level of abstraction hiding the details of the lower level.
\end{shaded}

\section{Introduction to Digital Logic}

@@

\section{Digital Storage Elements}

@@

\section{Computer Organization}

>>>>

\documentclass[11pt]{report}
\input{preamble}
\begin{document}
\thispagestyle{empty}
{\LARGE \bf ECE 253 Lecture Notes}\\
{\large Hei Shing Cheung}\\
Digital and Computer Systems, Fall 2025 \hfill ECE253\\
\\
The up-to-date version of this document can be found at \url{https://github.com/HaysonC/skulenotes}\\

\chapter{Digital Circuits that Compute, Store, and Control}
\begin{shaded}
\section*{Introduction}
\paragraph{Layers of Computation} In hardware, we have the following layers of abstraction:
\begin{itemize}
    \item Computation
    \item Adders 
    \item Logic Gates
    \item Transistors
    \item Silicon
\end{itemize}
In this course, we will focus on the first three layers, on top of the logic gate level.

\paragraph{Layer of abstraction} At this course, for the digital systems part, we would start from understanding logic gates, all the way to understanding computer architecture, with each level of abstraction hiding the details of the lower level.

@@

\paragraph{Factors Affecting Speed of Digital Circuits}
\begin{itemize}
    \item \textbf{Transistors and Electrons take time to switch.} A transistor (State of the Art) takes 2-3 picoseconds to switch. Gates takes 40 ps and an 8-bit adder takes 300 ps.
    \item \textbf{Wires take time to propagate signals.} Signals travel at approximately 2/3 the speed of light in a vacuum, which is about 200,000 kilometers per second in a typical silicon wire.
    \item \textbf{Capacitance} There would be RCL circuits formed by the wires and transistors, which would cause delay.
\end{itemize}

\end{shaded}

\section{Information Representation}

\subsection{Number Systems}
\begin{definition}[Number System]
    A number system is a way of representing numbers using a set of symbols (digits) and a base (radix). The base determines the number of unique digits that can be used in the number system.
\end{definition}

\paragraph{Common Number Systems} You should be familiar with the following number systems:
\begin{itemize}
    \item Decimal (Base 10): Digits 0-9
    \item Binary (Base 2): Digits 0-1
    \item Hexadecimal (Base 16): Digits 0-9, A-F
\end{itemize}

In computer systems, we use binary to represent information, and we would often use hexadecimal to represent binary numbers in a more compact way - a group of 4 bits (a nibble) can be represented by a single hexadecimal digit.

\begin{example}[Binary, Decimal, and Hexadecimal Numbers]
    Below is a table showing the conversion of binary numbers to decimal numbers, along with their hexadecimal representation.
    \begin{table}[h!]
        \centering
        \begin{tabular}{|c|c|c|}
            \hline
            Binary & Decimal & Hexadecimal \\
            \hline
            0000 & 0 & 0 \\
            0001 & 1 & 1 \\
            0010 & 2 & 2 \\
            0011 & 3 & 3 \\
            0100 & 4 & 4 \\
            0101 & 5 & 5 \\
            0110 & 6 & 6 \\
            0111 & 7 & 7 \\
            1000 & 8 & 8 \\
            1001 & 9 & 9 \\
            1010 & 10 & A \\
            1011 & 11 & B \\
            1100 & 12 & C \\
            1101 & 13 & D \\
            1110 & 14 & E \\
            1111 & 15 & F \\
            \hline
        \end{tabular}
        \caption{Binary to Decimal and Hexadecimal Conversion}
    \end{table}

\end{example}
\begin{example}[Decimal to Binary Conversion]
    To convert a decimal number to binary, we can use the method of successive division by 2. For example, to convert the decimal number 437 to binary:
    \begin{align*}
        437 \div 2 &= 218 & \text{remainder 1} \\
        218 \div 2 &= 109 & \text{remainder 0} \\
        109 \div 2 &= 54 & \text{remainder 1} \\
        54 \div 2 &= 27 & \text{remainder 0} \\
        27 \div 2 &= 13 & \text{remainder 1} \\
        13 \div 2 &= 6 & \text{remainder 1} \\
        6 \div 2 &= 3 & \text{remainder 0} \\
        3 \div 2 &= 1 & \text{remainder 1} \\
        1 \div 2 &= 0 & \text{remainder 1}
    \end{align*}
    Reading the remainders from bottom to top, we get the binary representation of 437
\end{example}

\begin{example}
    To convert (512000)$_{10}$ to binary, we recognize that $512000 = 2^9 \times 1000$. We know that $2^9 = 512$ and $1000_{10} = 1111101000_2$ (by method outlined above). Therefore, we can shift the binary representation of 1000 left by 9 bits to get the binary representation of 512000:
    \[(512000)_{10} = (1111101000000000000)_2\]
\end{example}

\paragraph{Note} An alternative method is to devide by powers of 2.

\paragraph{Binary Arithmetic} Binary arithmetic is similar to decimal arithmetic, but it only uses two digits (0 and 1). The 0 and 1 could be associated with a xor logic.
\section{Digital Storage Elements}

@@

\chapter{Computer Organization and Assembly Language}
\begin{shaded}
\textbf{What is Assembly Language?} We know that we can run C/C++ on any computer (Machine Agnostic), but how does the computer understand C/C++? The answer is the compiler that parse it to assembly through:

\begin{enumerate}
    \item \textbf{Front-end Parser:} The front-end parser would parse the C/C++ code into an intermediate representation (IR), which is a low-level representation of the code that is easier to optimize. The front-end parser would also perform optimizations on the IR, such as loop unrolling, inlining, and dead code elimination.
    \item \textbf{Back-end Parser:} The back-end parser would take the optimized IR and generate assembly code for a specific architectures.
\end{enumerate}

The assembly code is then assembled into machine code, which is a series of 0s and 1s that the computer can understand. The assembly would be specific to the architecture of the computer (machine dependent), which is why we have different assembly languages for different architectures (e.g., x86, RISC-V, ARM).
\end{shaded}
\section{Computer Organization}
